---
title: "Antibiotics"
author: "Khor Zhi Hong"
date: 2020-3-4
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
df_antibiotics
```

# Visualization
<!-- -------------------------------------------------- -->

### __q1__ Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE


df_antibiotics_1 <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin, streptomycin, neomycin), names_to = "antibiotic", values_to = "MIC") %>%
    mutate(gram_status = ifelse(gram == "positive", "Gram-positive", "Gram-negative"))


df_antibiotics_1 %>%
  ggplot(aes(x = bacteria, y = MIC, fill = antibiotic)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("penicillin" = "green", "streptomycin" = "blue", "neomycin" = "orange")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  facet_wrap(~ gram_status) +
  labs(
    title = "MIC of Antibiotics on Different Bacteria",
    x = "Bacteria",
    y = "MIC",
    fill = "Antibiotic")





```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE

df_antibiotics_2 <- df_antibiotics %>%
  mutate(gram_status = ifelse(gram == "positive", "Gram-positive", "Gram-negative")) %>%
  pivot_longer(
    cols = c(penicillin, 
    streptomycin, 
    neomycin), 
    names_to = "antibiotic", 
    values_to = "MIC"
    )

df_antibiotics_2 %>%
  ggplot(aes(x = bacteria, y = log10(MIC), color = gram_status)) +
  geom_point(aes(shape = antibiotic), size = 3) +
  scale_color_manual(values = c("Gram-positive" = "red", "Gram-negative" = "blue")) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
  title = "MIC of Antibiotics on Different Bacteria",
  x = "Bacteria",
  y = "log 10 MIC",
  color = "Gram Status",
  shape = "Antibiotic"
)


```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}

df_antibiotics_3 <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin, streptomycin, neomycin), names_to = "antibiotic", values_to = "MIC") %>%
  mutate(gram_status = ifelse(gram == "positive", "Gram-positive", "Gram-negative"))


df_antibiotics_3 %>%
  ggplot(aes(x = antibiotic, y = MIC, fill = antibiotic)) +
  geom_bar(stat = "identity", position = "dodge", width = 1) +
  scale_fill_manual(values = c("penicillin" = "darkorange", "streptomycin" = "forestgreen", "neomycin" = "dodgerblue")) +
  facet_wrap(~ bacteria, scales = "free_y") +
  labs(title = "MIC of Antibiotics on Different Bacteria",
       x = "Antibiotic",
       y = "MIC",
       fill = "Antibiotic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE

df_antibiotics_4 <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin, streptomycin, neomycin), names_to = "antibiotic", values_to = "MIC") %>%
  mutate(gram_status = ifelse(gram == "positive", "Gram-positive", "Gram-negative"))


df_antibiotics_4 %>%
  ggplot(aes(x = antibiotic, y = bacteria, size = MIC, color = gram_status)) +
  geom_point(alpha = 0.7) +
  scale_size_continuous(range = c(2, 11)) +
  scale_color_manual(values = c("Gram-positive" = "red", "Gram-negative" = "blue")) +
  labs(
  title = "MIC of Antibiotics on Different Bacteria",
  x = "Antibiotic",
  y = "Bacteria",
  size = "MIC",
  color = "Gram Status") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) 

```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE

df_antibiotics_5 <- df_antibiotics %>%
  pivot_longer(cols = c(penicillin, streptomycin, neomycin), names_to = "antibiotic", values_to = "MIC") %>%
  mutate(gram_status = ifelse(gram == "positive", "Gram-positive", "Gram-negative"))

df_antibiotics_5 %>%
  ggplot(aes(x = log10(MIC), y = bacteria, color = antibiotic)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("penicillin" = "orange", "streptomycin" = "green", "neomycin" = "blue")) +
  facet_wrap(~ gram_status, scales = "free_y") +
  labs(
  title = "MIC of Antibiotics on Different Bacteria",
  x = "log 10 MIC",
  y = "Bacteria",
  color = "Antibiotic") 


```

### __q2__ Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

*Observations*
- What is your response to the question above?
  - The effectiveness of the three antibiotics (penicillin, streptomycin, and neomycin) varies significantly depending on the  genera and Gram stain status. 
  
  -Penicillin tends to be less effective against Gram-neative bacteria, and especially ineffective against Aerobacter aerogenes, Klebsiella pneumoniae, Mycobacterium tuberculosis and Pseudomonas aeruginosa with MIC values ~800 and above.
  
  - streptomycin and neomycin have a broader range of effectiveness against both Gram-positive and Gram-negative bacteria, but have both have slight similar effect for most of the bacteria.
  
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - Visual 2, df_antibiotics_2: Dot Plot with shapes showing the type antibodies 
  
- Why?
  - df_antibiotics_3 is most effective to show which of antibiotics MIC is below 0.1 but is lacking the Gram stain categorization thus not chosen. The Dot Plot with shapes showing antibodies, df_antibiotics_2 is most effective out of all 5 for this analysis because it shows the MIC of all three antibiotics for each bacterium and distinguishes between Gram-positive and Gram-negative bacteria using color coding. The shapes allows to compare the effectiveness of antibiotics within and between Gram-positive and Gram-negative groups.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

*Observations*
- What is your response to the question above?
  - From googling, In 1886, S.pneumoniae was first known as pneumococcus due to its close association with pneumonia. In 1920, it was renamed as Diplococcus pneumoniae as it was usually observed in pairs. In 1974, the name was changed to Streptococcus pneumoniae due to its similarities in forming chains like other Streptococcus spp. S. pneumoniae can be distinguished from other species using rMLST.
  
- Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?
  - None 
- Why?
  - It is a historical and biological question rather than one that can be answered directly through visual data analysis. The visuals  are more suited for analyzing the effectiveness of antibiotics against bacteria and their Gram stain status, but not for reason why name changes.


# References
<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
